\documentclass[11pt]{article}

\usepackage[a4paper, margin=2.5cm]{geometry}
\usepackage{amsmath, amssymb, bm}
\usepackage{hyperref}

\title{Why is the residual $\bm{z}$ uncorrelated with $\bm{x}_1$?\\
       \large Supplementary note for AS4501 (Gaussian process derivation)}
\author{Francisco F\"orster Bur\'on}
\date{\today}

\begin{document}
\maketitle

\section*{Setting}

Let $\bm{x} = (\bm{x}_1, \bm{x}_2)^T$ be jointly Gaussian,
\[
  \bm{x} \sim \mathcal{N}\!\left(
    \begin{pmatrix} \bm{\mu}_1 \\ \bm{\mu}_2 \end{pmatrix},
    \;
    \Sigma = \begin{pmatrix} \Sigma_{11} & \Sigma_{12} \\ \Sigma_{21} & \Sigma_{22} \end{pmatrix}
  \right),
  \qquad \Sigma_{12} = \Sigma_{21}^T.
\]

In the Gaussian-process posterior derivation we used the following
identity, here stated for completeness:
\[
  \bm{x}_2 \mid \bm{x}_1 \;\sim\; \mathcal{N}\!\Bigl(
     \bm{\mu}_2 + \Sigma_{21}\Sigma_{11}^{-1}(\bm{x}_1 - \bm{\mu}_1), \;\;
     \Sigma_{22} - \Sigma_{21}\Sigma_{11}^{-1}\Sigma_{12}
  \Bigr).
\]
The proof of this identity hinges on the claim that the
\emph{residual}
\begin{equation}\label{eq:zdef}
  \bm{z} \;=\; \bm{x}_2 \;-\; \Sigma_{21}\Sigma_{11}^{-1}\,\bm{x}_1
\end{equation}
is uncorrelated with $\bm{x}_1$. The purpose of this note is to spell
out \emph{why} that is true.

\section*{Algebraic verification}

Using bilinearity of covariance,
\begin{align}
  \mathrm{Cov}(\bm{z}, \bm{x}_1)
  &= \mathrm{Cov}\bigl(\bm{x}_2 - \Sigma_{21}\Sigma_{11}^{-1}\bm{x}_1,\; \bm{x}_1\bigr) \\
  &= \mathrm{Cov}(\bm{x}_2, \bm{x}_1) \;-\; \Sigma_{21}\Sigma_{11}^{-1}\,\mathrm{Cov}(\bm{x}_1, \bm{x}_1).
\end{align}

Substituting the known block-covariance entries
$\mathrm{Cov}(\bm{x}_2, \bm{x}_1) = \Sigma_{21}$ and
$\mathrm{Cov}(\bm{x}_1, \bm{x}_1) = \Sigma_{11}$:
\begin{align}
  \mathrm{Cov}(\bm{z}, \bm{x}_1)
  &= \Sigma_{21} - \Sigma_{21}\Sigma_{11}^{-1}\Sigma_{11}
   = \Sigma_{21} - \Sigma_{21}
   = \bm{0}.
\end{align}

The algebra is short, but the choice of coefficient
$\Sigma_{21}\Sigma_{11}^{-1}$ looked unmotivated. The next two sections
explain why this is precisely the right coefficient.

\section*{Where the coefficient comes from}

Look for the matrix $A$ that makes
$\mathrm{Cov}(\bm{x}_2 - A\bm{x}_1,\; \bm{x}_1) = 0$. Expanding,
\[
  \mathrm{Cov}(\bm{x}_2 - A\bm{x}_1,\; \bm{x}_1)
  = \Sigma_{21} - A\,\Sigma_{11} = \bm{0}
  \quad\Longleftrightarrow\quad
  A = \Sigma_{21}\,\Sigma_{11}^{-1}.
\]
There is a \emph{unique} linear combination of $\bm{x}_1$ that, when
subtracted from $\bm{x}_2$, removes all linear dependence. That is the
content of equation~\eqref{eq:zdef}.

\section*{Geometric intuition}

Consider the inner product on random variables given by
$\langle u, v\rangle := \mathrm{Cov}(u, v)$. With this inner product,
\emph{uncorrelated} means \emph{orthogonal}.

The vector $A\bm{x}_1 = \Sigma_{21}\Sigma_{11}^{-1}\bm{x}_1$ is the
\emph{orthogonal projection} of $\bm{x}_2$ onto the linear span of the
components of $\bm{x}_1$. The residual $\bm{z}$ is, by definition, the
component of $\bm{x}_2$ perpendicular to that span. The Pythagorean
decomposition reads
\[
  \bm{x}_2 \;=\; \underbrace{\Sigma_{21}\Sigma_{11}^{-1}\,\bm{x}_1}_{\text{projection}} \;+\; \underbrace{\bm{z}}_{\perp \,\bm{x}_1}.
\]
The same fact appears in ordinary least squares: residuals are
orthogonal to the regressors.

\section*{Connection to linear regression}

Given centred $\bm{x}_1$ and $\bm{x}_2$, the least-squares regression
coefficient
\[
  A^\star \;=\; \arg\min_A \mathbb{E}\bigl[\|\bm{x}_2 - A\bm{x}_1\|^2\bigr]
\]
has the closed form $A^\star = \Sigma_{21}\Sigma_{11}^{-1}$. So the
construction of $\bm{z}$ is exactly: \emph{subtract the best linear
predictor of $\bm{x}_2$ given $\bm{x}_1$.} Whatever is left is, by the
defining optimality, uncorrelated with the regressor. This is the
``magic'' that makes the Gaussian conditioning identity work.

\section*{Why uncorrelated $\Rightarrow$ independent (Gaussian only)}

For \emph{arbitrary} distributions, uncorrelated does not imply
independent. For \emph{jointly Gaussian} variables it does. Since
$\bm{z}$ is an affine function of $\bm{x} = (\bm{x}_1, \bm{x}_2)$, the
pair $(\bm{z}, \bm{x}_1)$ is again jointly Gaussian. A jointly
Gaussian pair with zero covariance has joint density that factors into
a product of marginals, hence the variables are independent. This is
what lets us write $\bm{z} \mid \bm{x}_1 = \bm{z}$ in the GP proof and
read off the conditional moments by substituting
$\bm{x}_2 = \bm{z} + \Sigma_{21}\Sigma_{11}^{-1}\bm{x}_1$:
\begin{align}
  \mathbb{E}[\bm{x}_2 \mid \bm{x}_1]   &= \bm{\mu}_2 + \Sigma_{21}\Sigma_{11}^{-1}\,(\bm{x}_1 - \bm{\mu}_1), \\
  \mathrm{Var}(\bm{x}_2 \mid \bm{x}_1) &= \mathrm{Var}(\bm{z}) \;=\; \Sigma_{22} - \Sigma_{21}\Sigma_{11}^{-1}\Sigma_{12}.
\end{align}

\end{document}
